---
title: "Movie Recommendation System"
author: "Nathan Siefken"
date: "1/23/2019"
output: pdf_document
---
 
  <p>
##Introduction:
  
  A recommender system is an information filtering system that tries to predict the ratings a user would give a product. Some of the most dominate companies, like Amazon and Netflix not only have excellent products they provide a personalized experience as an extension of their product. They allow users to rate their products, then they use this information to predict an individual customer's rating for any particular product. They can to a degree of certainty anticipate the needs of their customers. 

In the case of Netflix, a low score is a one star and a great movie would be rated five stars. Netflix thought predicting customer ratings was so crucial to their business they hosted a competition in which, they awarded $1,000,000 to anyone who could improve their own recommendation system by 10 percent or more. We will cover one of the methods used by the winners, matrix factorization.   

*Note: I decided to leave the code in the report in case a reviewer wanted to see it here in addition to my script. In a work setting I would probably leave it out. *

```{r, warning=FALSE}
# This code chunk simply makes sure that all the libraries used here are installed. 
packages <- c("knitr","dplyr",  "tidyr", "caret", "ggplot2")
if ( length(missing_pkgs <- setdiff(packages, rownames(installed.packages()))) > 0) {
message("Installing missing package(s): ", paste(missing_pkgs, collapse = ", "))
install.packages(missing_pkgs)
}
```

```{r, warning=FALSE, message=FALSE}
# These are the libraries needed for my report
library(knitr)
library(tidyr)
library(dplyr)
library(ggplot2)

```
*The Data :*

In this project, I will be creating a movie recommendation system using the [ 10M MovieLens Dataset](https://grouplens.org/datasets/movielens/10m/). Dataset. This data set includes 10 million ratings on 10 thousand movies by 72 thousand users. 10 % percent of this data was partitioned to create a validation set that will be used as a final test by HarvardX to measure how well our algorithm predicts unseen data. The rest of the data is the data set, edx, used for this project.  

I have further partitioned this into my training set, labeled my_train, and a test set, labeled my_test, to test my algorithm before submission.  

```{r,echo=FALSE,comment=NA,results='hide'}
file_location <- "my_train.csv"
my_train <- read.csv(file_location, header = TRUE)
```

```{r,echo=FALSE,comment=NA,results='hide'}
file_location <- "my_test.csv"
my_test <- read.csv(file_location, header = TRUE)
```

<p>
## Methods Section: 

Matrix factorization takes information from a sparse matrix like ratings from users on different movies and fills in the empty values with a rating for each movie and each user. The ability to take global input from the data and produce meaningful output is extreamly powerful and we will go through this process in this report.

In comparison, the Neighborhood approach is popular, but it can be very localized. For example if a user likes The Matrix Reloaded, it will suggest they will like The Matrix. It still can be very usefull but at times the results do not seem that insightful. 

\pagebreak

*Measure the quality of the algoriths*

Netflix used the used Root Mean Squared Errors (RMSE) to measure the quality of the algorithms in the challenge. The winning team Bellcore Pragmatic Chaos won the challenge with a RMSE of .8558 which was an improvement of 10.05 of Netflix's recommender system. The goal of this report is to beat the results of Bellcore Pragmatic Chaos and have a lower RMSE than .85558.

We define $y_{u,i}$ as the rating for movie $\imath$ and user ${u}$ and denote our prediction as   $\hat{y}_{i,u}$
  <br>
  \[
    \makebox[\linewidth]{$RMSE = \sqrt{\frac{1}{n}\Sigma_{i=1}^{n}{\Big(\frac{d_i -f_i}{\sigma_i}\Big)^2}}$}
    \]
<br>
  As described in [Wikipedia](https://en.wikipedia.org/wiki/Root-mean-square_deviation)
<p>
  >Root Mean Square Error (RMSE) is a frequently used measure of the differences between values (sample or population values) predicted by a model or an estimator and the values observed.

Since it is a measurement of error, the smaller the RMSE the better. To better understand how to interpret the RMSE for our algorithm, if we receive a RMSE of 1, means that our prediction was off by an average of 1 star. 

We should start by building a function  that will calculate the RMSE for actual values (true_ratings) from our test set to their corresponding predictors from our models:
  
```{r}
RMSE <- function(true_ratings, predicted_ratings){
  sqrt(mean((true_ratings - predicted_ratings)^2))
}
```

<p>
  <p>
  
*Exploration of the data:*

```{r }
dim(my_train)

```

The training set has 7,200,043 ratings with the following 6 variables. 

```{r}
colnames(my_train)
```
This code will show the number of unique movies and users.
```{r}

my_train %>%
  summarize(Users= n_distinct(userId),
            Movies = n_distinct(movieId) )
```

My training data set has 69,878 users have rated 10,641 different movies for a grand total of 7,200,043 observations. For each of the observations we have the following variables: userId, movieId, rating, timestamp, title, and genres.

\pagebreak

Below shows a histogram the number of times a few movies have been rated. The movies that have been rated more than 10,000 times is going to carry much more weight in a calculation than the movies on the left side of the graph that have been rated only 10 times or fewer. This is known as the movie bias or effect. 

```{r}
my_train %>% 
  count(movieId) %>% 
  ggplot(aes(n)) + 
  geom_histogram(bins = 30, color = "skyblue") + 
  scale_x_log10() + 
  ggtitle("Movies")

```

\pagebreak

We can see there appears  to be a similar bias among users. Some users are extremely active while some have only reviewed a few movies.

```{r}
my_train %>%
  count(userId) %>% 
  ggplot(aes(n)) + 
  geom_histogram(bins = 30, color = "orange") + 
  scale_x_log10() + 
  ggtitle("Users")
```

\pagebreak
<p>

To get a better perspective of our data, let's see the top 10 genres for the movies in our data set.

```{r}
genreCount <-my_train %>% separate_rows(genres, sep = "\\|") %>%
group_by(genres) %>%
summarize(count = n()) %>%
arrange(desc(count)) %>%
head(n=10)
genreCount %>% knitr::kable()
```

Drama and Comedy are the most likely to be rated. 

The following are the top 10 most rated the movies.
```{r}
Top10 <- my_train %>% group_by(movieId, title) %>%
summarize(count = n()) %>%
arrange(desc(count)) %>% head(n=10)
Top10 %>% knitr::kable()
```

\pagebreak

<p>

Let's explore to see the most common rating given by the users.

```{r}
ratings <- my_train %>% group_by(rating) %>% summarize(count = n()) %>% 
  mutate(percent = count/nrow(my_train)*100) %>%
  top_n(5) %>%
  arrange(desc(count))  
ratings %>% knitr::kable()
```

It is important to note that the users rated movies 4 stars over a quarter of the time. 

Now that we have taken a look at the data, we should start building our recommendation machine

\pagebreak

#Baseline

We will start with our baseline, the most basic recommendation system possible. This is the average for all users across all movies and use this average to predict our ratings.

\[
  \makebox[\linewidth]{${y}_{u,i}=\mu+\varepsilon_{u,i}$}
  \]

```{r}
mu_hat <- mean(my_train$rating)

mu_hat
```

Now that we have our $\hat{\mu}$ we can determine RMSE for our baseline method. 


```{r}
baseline_rmse <- RMSE(my_test$rating, mu_hat)

baseline_rmse
```

We are getting a RMSE of about 1.06. Our prediction are on average one star off the actual data. Not all that accurate. However, for this model, this is the lowest RMSE we can have. To demonstrate we will use 4, the most common rating given by the users. We can see below that our RMSE is higher than when we use $\mu$.

```{r}
predictions <- rep(4, nrow(my_test))
RMSE(my_test$rating, predictions)
```

The following code will keep track with all of our test with this matrix.

```{r}
rmse_results <- data_frame(method = "Baseline", RMSE = baseline_rmse)

rmse_results %>% knitr::kable()
```

This model utilizes $\mu$ but does not take into account the movie or user bias, so there is opportunity for improvement.

\pagebreak

# Modeling Movie Effect

To improve on our model, we should consider the movie bias $i$, as we mentioned the uneven amount of times movies are rated. The histogram below support this claim with data.

Just to better explain the graph. $\mu$ in our case is 3.5, therefore a $b_i$ of -3 is our lowest rating of .5 and $b_i$ of 1.5 would b a five-star  rating. 

```{r}
mu <- mean(my_train$rating) 
movie_avgs <- my_train %>% 
  group_by(movieId) %>% 
  summarize(b_i = mean(rating - mu))
```

```{r}
movie_avgs %>% qplot(b_i, geom ="histogram", bins = 10, data = ., color = I("skyblue"))
```

*Note: The lm function is excellent to use but we have too many values in this data set and it will take too long to run, possibly error out due to lack of computer memory.*
  \[
    \makebox[\linewidth]{${y}_{u,i}=\mu+b_i+\varepsilon_{u,i}$}
    \]
In this situation we can estimate the movie bias $\hat{b}_\imath$ is just the average of ${y}_{u,i}-\mu$ for each movie $\imath$. So we can compute them his way (for computational ease in r we will drop the hat notation)

Let's see how our model did with the following equation. We will name it the "Movie Effect Model"
\[
  \makebox[\linewidth]{${\hat{y}}_{u,i}=\hat{\mu}+\hat{b}_i$}
  \]


```{r}
predicted_ratings <- mu_hat + my_test %>% 
  left_join(movie_avgs, by='movieId') %>%
  .$b_i

model_1_rmse <- RMSE(predicted_ratings, my_test$rating)
rmse_results <- bind_rows(rmse_results,
                          data_frame(method="Movie Effect Model",  
                                     RMSE = model_1_rmse ))
rmse_results %>% knitr::kable()
```

There is a modest improvement when we include the movie bias to our baseline model.

\pagebreak

#Movie and User Effects Model

There is further opportunity to improve our model by including the users bias also known as the user effect $b_u$. 

We can visualize in the difference in the nature of the users rating habits in the histogram below.

```{r}
my_train %>% 
  group_by(userId) %>% 
  summarize(b_u = mean(rating)) %>% 
  filter(n()>=100) %>%
  ggplot(aes(b_u)) + 
  geom_histogram(bins = 30, color = "orange")
```

As you can see the users on the left on average rate movies poorly and the users on the right on average enjoy every movie. Taking this into account should improve our model.

```{r}
user_avgs <- my_train %>% 
  left_join(movie_avgs, by='movieId') %>%
  group_by(userId) %>%
  summarize(b_u = mean(rating - mu_hat - b_i))
```

Below we can see how our base model has improved when we include both the movie and user bias to our baseline.  
  
```{r}
predicted_ratings <- my_test %>% 
  left_join(movie_avgs, by='movieId') %>%
  left_join(user_avgs, by='userId') %>%
  mutate(pred = mu_hat + b_i + b_u) %>%
  .$pred


model_2_rmse <- RMSE(predicted_ratings, my_test$rating)
rmse_results <- bind_rows(rmse_results,
                          data_frame(method="Movie + User Effects Model",  
                                     RMSE = model_2_rmse ))
rmse_results %>% knitr::kable()
```

This is a great improvement from our baseline and is clse to what the winners of the Netflix achieved. I think it is worth taking a look at our results to see if there is an additional opportunity for improvement.

\pagebreak

# Regularization

First, let’s look at our largest errors in our *Movie Effect Model* to see if we can understand what happened. Here is a list of the largest errors.
```{r}
my_test %>% 
  left_join(movie_avgs, by='movieId') %>%
  mutate(residual = rating - (mu_hat + b_i)) %>%
  arrange(desc(abs(residual))) %>% 
  select(title,  residual) %>% slice(1:10) %>% knitr::kable()
```

The first thing I notice is that these movies repeat, so the error repeats many times in our matrix. We will run the following code that will eliminate duplicated movies errors to better understand unique mistakes.

```{r}
movie_titles <- my_train %>% 
  select(movieId, title) %>%
  distinct()
```

The following code shows will produce the 10 best movies according to our model.

```{r}
movie_avgs %>% left_join(movie_titles, by="movieId") %>%
  arrange(desc(b_i)) %>% 
  select(title, b_i) %>% 
  slice(1:10) %>%  
  knitr::kable()
```

This is interesting that all of these movies are unrecognizable. It doesn't seem fitting for the best movies rated by users. 

These are the 10 worst movies according to our model.
```{r}
movie_avgs %>% left_join(movie_titles, by="movieId") %>%
  arrange(b_i) %>% 
  select(title, b_i) %>% 
  slice(1:10) %>%  
  knitr::kable()
```

Now I would be curious to know how many times these movies have been rated.

The following counts how many times our top 10 movies have been rated in our training set. 
```{r}
my_train %>% count(movieId) %>% 
  left_join(movie_avgs) %>%
  left_join(movie_titles, by="movieId") %>%
  arrange(desc(b_i)) %>% 
  select(title, b_i, n) %>% 
  slice(1:10) %>% 
  knitr::kable()
```

The following counts how many times our top 10 worst movies have been rated in our training set.  

```{r}
my_train %>% count(movieId) %>% 
  left_join(movie_avgs) %>%
  left_join(movie_titles, by="movieId") %>%
  arrange(b_i) %>% 
  select(title, b_i, n) %>% 
  slice(1:10) %>% 
  knitr::kable()
```

Just as we suspected these movies that raked at the top and bottom were only rated a few times. These movies do not have enough ratings to reflect a true rating of the movie or perhaps they do reflect the true rating. The point is we do not know, and we should be skeptical of these results.

To optimize \(b_i\) the following equation can be used. This is the least squares and the second term is the penalty which gets lareger as \(b_i\) gets larger.

\[\frac{1}{N} \sum_{u,i} \left(y_{u,i} - \mu - b_i\right)^2 + \lambda \sum_{i} b_i^2\]

This is equation can be reduced to the following equation.

\[ \hat{b}_i(\lambda) = \frac{1}{\lambda + n_i} \sum_{u=1}^{n_i} \left(Y_{u,i} - \hat{\mu}\right) \]

We can use the method Regularization which adds a penalty lambda (in this case 2.25) to penalizes movies with large estimates from a small sample size, in our case only rated a few times.

```{r}
lambda <- 2.25
mu <- mean(my_train$rating)
movie_reg_avgs <- my_train %>% 
  group_by(movieId) %>% 
  summarize(b_i = sum(rating - mu)/(n()+lambda), n_i = n()) 
```


The next code creates a graph that shows how the estimates shrink with the penalty.

```{r}
data_frame(original = movie_avgs$b_i, 
           regularlized = movie_reg_avgs$b_i, 
           n = movie_reg_avgs$n_i) %>%
  ggplot(aes(original, regularlized, size=sqrt(n))) + 
  geom_point(shape=1, alpha=0.5)
```

\pagebreak


These are the 10 best rated movies after regularization. This makes a lot more sense because these are recognizable and have been rated thousands of times.

```{r}
my_train %>%
  count(movieId) %>% 
  left_join(movie_reg_avgs) %>%
  left_join(movie_titles, by="movieId") %>%
  arrange(desc(b_i)) %>% 
  select(title, b_i, n) %>% 
  slice(1:10) %>% 
  knitr::kable()
```

These are the 10 worst rated movies after regularization. Even though they may not be recognizable, we can trust their rating as they have been rated several times.

```{r}
my_train %>%
  count(movieId) %>% 
  left_join(movie_reg_avgs) %>%
  left_join(movie_titles, by="movieId") %>%
  arrange(b_i) %>% 
  select(title, b_i, n) %>% 
  slice(1:10) %>% 
  knitr::kable()
```

This code measures how well our predictions preformed on the my_test set and then adds the RMSE to our results table.

```{r}
predicted_ratings <- my_test%>% 
  left_join(movie_reg_avgs, by='movieId') %>%
  mutate(pred = mu_hat + b_i) %>%
  .$pred

model_3_rmse <- RMSE(predicted_ratings, my_test$rating)
rmse_results <- bind_rows(rmse_results,
                          data_frame(method="Regularized Movie Effect Model",  
                                     RMSE = model_3_rmse ))
rmse_results %>% knitr::kable()
```

The result is slightly better than the results from the Movie Effect Model without Regularization. However now when you look at the data it makes more sense. The best movies are blockbusters and the worse movies are movies that have at least been rated enough for us to trust the rating.

\pagebreak


*Picking a penalty*

The following code shows how use cross validation to choose lambda. It will essentially find the RMSE for different penalties (from 0 to 10 at a sequence of .25) applied to the movie bias. We then can easily find the lambda that will produce the lowest RMSE.

We then can visualize the results of the different lambdas on a graph and we print out the \(\lambda\) that produces the smallest RMSE for \(b_i\). 

```{r}
lambdas <- seq(0, 10, 0.25)

mu <- mean(my_train$rating)
just_the_sum <- my_train %>% 
  group_by(movieId) %>% 
  summarize(s = sum(rating - mu), n_i = n())

rmses <- sapply(lambdas, function(l){
  predicted_ratings <- my_test %>% 
    left_join(just_the_sum, by='movieId') %>% 
    mutate(b_i = s/(n_i+l)) %>%
    mutate(pred = mu + b_i) %>%
    .$pred
  return(RMSE(predicted_ratings, my_test$rating))
})
qplot(lambdas, rmses)  
lambdas[which.min(rmses)]
```

We will repeat the step above and apply it to the user and movie bias to find the \(\lambda\)  that gives us the smallest RMSE.

\[ \frac{1}{N} \sum_{u,i} \left(y_{i,u} - \mu - b_i - b_u \right)^2 + \lambda \left(\sum_{i} b_i^2 + \sum_{u} b_u^2\right) \]

Again, we can visualize the results of the different lambdas on a graph and then print out the \(\lambda\) that produces the smallest RMSE for \(b_i\). 

```{r}
lambdas <- seq(0, 10, 0.25)

rmses <- sapply(lambdas, function(l){
  
  mu <- mean(my_train$rating)
  
  b_i <- my_train %>% 
    group_by(movieId) %>%
    summarize(b_i = sum(rating - mu)/(n()+l))
  
  b_u <- my_train %>% 
    left_join(b_i, by="movieId") %>%
    group_by(userId) %>%
    summarize(b_u = sum(rating - b_i - mu)/(n()+l))
  
  predicted_ratings <- 
    my_test %>% 
    left_join(b_i, by = "movieId") %>%
    left_join(b_u, by = "userId") %>%
    mutate(pred = mu + b_i + b_u) %>%
    .$pred
  
  return(RMSE(predicted_ratings, my_test$rating))
})

qplot(lambdas, rmses) 
```

This following is the \(\lambda\) for the lowest RMSE

```{r}
lambda <- lambdas[which.min(rmses)]
lambda
```

Regularization on the Movie + User Effect Model had the best results. 

```{r}
rmse_results <- bind_rows(rmse_results,
                          data_frame(method="Regularized Movie + User Effect Model",  
                                     RMSE = min(rmses)))
rmse_results %>% knitr::kable()
```

\pagebreak

<p>
#Conclusion: 
<p>
We have effectivly used machine learning methods to create a movie recommendation system. The methods that we used did not quite beat the winners of the $1 million Netfilx challange but we did get close.  We found that when we take into account both the user and the movie effect or bias we notice the largest improvement in RMSE. Also when we apply Regularization we do see a slight improvement on the Movie Effect Model and User + Movie Effect Model. However, the improvement is espcially effective on movies with small sample sizes and allows us to produce results that make sense. 


